\documentclass[final,5p]{elsarticle}
% \documentclass[preprint,onecolumn]{elsarticle}

% ============== Packages ==============
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{amsthm} % Added for proof and proposition environments
\newtheorem{proposition}{Proposition} % Define proposition environment
\usepackage{algorithm}
\usepackage{algpseudocode}  % Better algorithm support
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{multirow}
% \usepackage{cite}  % Commented out: conflicts with natbib loaded by elsarticle
\usepackage{subcaption}
\usepackage{pifont}  % For \ding symbols
\usepackage{tikz}    % For diagrams
\usetikzlibrary{shapes,arrows,positioning,fit,backgrounds}
\usepackage[colorlinks=true,linkcolor=blue]{hyperref}


\bibliographystyle{elsarticle-num}
% \bibliographystyle{elsarticle-harv}
% \bibliographystyle{elsarticle}
% \biboptions{authoryear}
% \setcitestyle{authoryear,round} % switch to ``Ren et al. (2019)'' style

% ============== Journal Options ==============
\journal{Information Sciences}

% ============== Custom Commands ==============
\newcommand{\ie}{\textit{i.e.}}
\newcommand{\eg}{\textit{e.g.}}
\newcommand{\etal}{\textit{et al.}}
\newcommand{\wrt}{w.r.t.}
\newcommand{\fake}[1]{\textcolor{red}{[FAKE: #1]}}

\newcommand{\ours}{EATA}
\newcommand{\ourslong}{Explainable Algorithmic Trading Agent}
\newcommand{\Yin}[1]{\textcolor{red}{#1}}
\newcommand{\RA}[1]{\textcolor{black}{#1}}
\newcommand{\RB}[1]{\textcolor{black}{#1}}
\newcommand{\RC}[1]{\textcolor{black}{#1}}
\newcommand{\hidden}[1]{%
    \textcolor{white}{\fontsize{0.1pt}{0.1pt}\selectfont #1}%
}
% ============== Colors ==============
\definecolor{eataBlue}{HTML}{4169E1}
\definecolor{profitGreen}{HTML}{50C878}
\definecolor{riskRed}{HTML}{DC143C}
\definecolor{baselineGray}{HTML}{808080}

% ============== Document ==============
\begin{document}

\begin{frontmatter}

% \title{\ours: \ourslong}
\title{Bridging Tail-Risk Objectives and Symbolic Search for Transparent Trading}

% %%% Blinded: authors and affiliations commented out for double-anonymized review
\author [a]{Jingtong Zhang}
\author [b]{Xiaoang Yang}
\author [c]{Yinqi Shi}
\author [d]{Yin Tang\corref{cor2}} %
\ead{Corresponding author: Yin Tang (Email: ytang@jnu.edu.cn).}
% \ead[url]{https://orcid.org/0000-0001-7693-7543}
% Author affiliation
\affiliation[a]{organization={University of New South Wales},
            city={Sydney},
            postcode={2052}, 
            country={Australia}}
\affiliation[b]{organization={School of Cybersecurity, Jinan University},
            % addressline={}, 
            city={GuangZhou},
            postcode={511443}, 
            country={China}}
\affiliation[c]{organization={Jinan University -- University of Birmingham Joint Institute, Jinan University},
            % addressline={}, 
            city={GuangZhou},
            postcode={511443}, 
            country={China}}
\affiliation[d]{organization={School of Management, Jinan University},%Department and Organization
            % addressline={}, 
            city={GuangZhou},
            postcode={510632}, 
            country={China}}


\begin{abstract}
Profit-seeking yet interpretable trading remains elusive: deep neural predictors excel at accuracy but violate regulatory transparency, whereas symbolic regression yields readable formulas but optimizes proxy objectives misaligned with economic utility. We present \ours~(\ourslong), a neural-guided symbolic regression framework that aligns objective, hypothesis space, and training signal with trading profitability. \ours{} couples a three-headed policy--value--profit network (PVPNet) with Monte Carlo Tree Search so that accuracy rewards $r^{\text{acc}}$ stabilize search while a Wasserstein-derived profit reward $r^{\text{rl}}$ captures asymmetric tail risk. A finance-grounded, decoupled grammar with bounded module libraries encodes moving-average, momentum, RSI, volatility, and conditional operators to curb hypothesis-space inflation, while sliding-window inheritance and quantile-consensus signals enable temporal adaptation. Across representative S\&P 500 constituents from 2020--2024, \ours{} delivers annualized return 17.6\%, Sharpe ratio 0.54, and maximum drawdown 34.0\%, achieving competitive performance with 7 baselines spanning technical indicators, gradient boosting, transformers, and FinRL agents while maintaining full interpretability; transaction-cost and Pareto-frontier analyses confirm robustness. Ablation and objective studies show that removing Wasserstein rewards, neural guidance, or grammar augmentation severely degrades performance, demonstrating that profit-aware symbolic regression can achieve interpretable trading strategies. Reported metrics are point estimates; comprehensive block-bootstrap, Reality Check, SPA, and deflated-Sharpe analyses are left to future work.
\end{abstract}

\begin{keyword}
Algorithmic Trading, Symbolic Regression, Wasserstein Distance, Explainable AI, Optimal Transport, Risk-Sensitive Learning
\end{keyword}

\end{frontmatter}

% \maketitle

% ============== Include Sections ==============
\input{1.intro}
\input{2.related}
\input{3.method}
\input{4.experiments}
\input{5.conclusion}
\section*{Acknowledgment}
This work was supported by the National Natural Science Foundation of China (No.62272198) and by the Research and Application of Expressway ``1+N'' Electromechanical O\&M Brain Based on Large Model Agents.

\textbf{Author Contribution}
Jingtong Zhang: Conceptualization, Methodology, and Writing Original draft preparation; 
Xiaoang Yang: Software, Data curation, Visualization; 
Yin Tang: Supervision, Writing- Reviewing and Editing, Management.

\textbf{Declaration of generative AI and AI-assisted technologies in the writing process} During the preparation of this work the authors used AI only for proof-reading. After using this tool/service, the authors reviewed and edited the content as needed and take full responsibility for the content of the publication.

% ============== Bibliography ==============
\bibliography{references}

% ============== Appendix ==============
\input{appendix}

\end{document}
